---
title: "Causal Language Analysis Report"
subtitle: "Analysis of Causal vs. Noncausal Terms in Research Document"
author: "Your Name"
date: today
date-format: long
format:
  html:
    code-fold: true
    code-tools: true
    toc: true
    toc-depth: 3
    number-sections: true
    theme: cosmo
    embed-resources: true
    df-print: paged
execute:
  echo: true
  message: false
  warning: false
---

```{r}
#| label: setup
#| include: false

# Set options
base::options(scipen = 999)

# Set seed for reproducibility
base::set.seed(123)
```

# Document Information

This report analyzes the frequency of causal and noncausal language in research documents, providing insights into the rhetorical approach used by the authors.

```{r}
#| label: load-libraries
#| message: false

# Load necessary libraries - commented out, using namespace syntax instead
# library(tidyverse)
# library(pdftools)
# library(rjson)
# library(jsonlite)
# library(flextable)
# library(ggplot2)
```

# Configuration

```{r}
#| label: configuration

# Define the path to your PDF file
pdf_input_path <- "pdf input/Lorentzen-The Relationship Between School Board Governance Behaviors and St2.pdf"

# Define paths to glossary files
causal_glossary_path <- 'glossaries/glossary_causal_terms.json'
noncausal_glossary_path <- 'glossaries/glossary_noncausal_terms.json'
```

# Data Extraction

## Read PDF Content

```{r}
#| label: read-pdf

# Read the text content of the PDF file
pdf_text <- pdftools::pdf_text(pdf_input_path) |>
  base::tolower()  # Convert text to lowercase for case-insensitive search

# Display first few characters
base::cat("PDF successfully loaded.\n")
base::cat("Number of pages:", base::length(pdf_text), "\n")
base::cat("Sample text (first 200 characters):\n")
base::cat(base::substr(pdf_text[1], 1, 200), "...\n")
```

## Load Search Term Glossaries

```{r}
#| label: load-glossaries

# Load the causal search terms from the JSON file
causal_terms_glossary <- jsonlite::fromJSON(causal_glossary_path)

# Load the noncausal search terms from the JSON file
noncausal_terms_glossary <- jsonlite::fromJSON(noncausal_glossary_path)

# Display counts
base::cat("Causal terms in glossary:", base::length(causal_terms_glossary), "\n")
base::cat("Noncausal terms in glossary:", base::length(noncausal_terms_glossary), "\n")
```

# Term Frequency Analysis

## Define Counting Function

```{r}
#| label: counting-function

# Function to count occurrences of whole words in text
count_terms_in_text <- function(term) {
  pattern <- base::paste0('\\b', term, '\\b')
  base::sum(stringr::str_count(pdf_text, stringr::regex(pattern, ignore_case = TRUE)))
}
```

## Count Causal Terms

```{r}
#| label: count-causal-terms

# Map the counting function over causal search terms
causal_term_counts <- purrr::map_int(causal_terms_glossary, count_terms_in_text)

# Combine causal search terms and their counts into a tibble
causal_term_counts_df <- tibble::tibble(
  term = causal_terms_glossary, 
  count = causal_term_counts
) |>
  dplyr::arrange(dplyr::desc(count))

# Display summary
base::cat("Total causal term occurrences:", base::sum(causal_term_counts_df$count), "\n")
base::cat("Unique causal terms found:", base::sum(causal_term_counts_df$count > 0), "\n")
```

## Count Noncausal Terms

```{r}
#| label: count-noncausal-terms

# Map the counting function over noncausal search terms
noncausal_term_counts <- purrr::map_int(noncausal_terms_glossary, count_terms_in_text)

# Combine noncausal search terms and their counts into a tibble
noncausal_term_counts_df <- tibble::tibble(
  term = noncausal_terms_glossary, 
  count = noncausal_term_counts
) |>
  dplyr::arrange(dplyr::desc(count))

# Display summary
base::cat("Total noncausal term occurrences:", base::sum(noncausal_term_counts_df$count), "\n")
base::cat("Unique noncausal terms found:", base::sum(noncausal_term_counts_df$count > 0), "\n")
```

# Overall Statistics

## Calculate Document-Wide Percentages

```{r}
#| label: calculate-percentages

# Convert pdf_text to a single string
pdf_text_combined <- base::tolower(base::paste(pdf_text, collapse = " "))

# Calculate total words
total_words <- stringr::str_count(pdf_text_combined, stringr::boundary("word"))

# Calculate occurrences of each causal search term and sum them
causal_terms_total_occurrences <- base::sum(base::sapply(causal_terms_glossary, function(term) {
  pattern <- base::paste0('\\b', term, '\\b')
  base::sum(stringr::str_count(pdf_text_combined, stringr::regex(pattern, ignore_case = TRUE)))
}))

# Calculate occurrences of each noncausal search term and sum them
noncausal_terms_total_occurrences <- base::sum(base::sapply(noncausal_terms_glossary, function(term) {
  pattern <- base::paste0('\\b', term, '\\b')
  base::sum(stringr::str_count(pdf_text_combined, stringr::regex(pattern, ignore_case = TRUE)))
}))

# Calculate the percentage of causal terms in document
causal_percentage <- (causal_terms_total_occurrences / total_words) * 100

# Calculate the percentage of noncausal terms in document
noncausal_percentage <- (noncausal_terms_total_occurrences / total_words) * 100
```

## Summary Statistics Table

```{r}
#| label: summary-table
#| tbl-cap: "Document-Wide Language Analysis Summary"

# Create summary data frame
summary_stats <- base::data.frame(
  Metric = base::c(
    "Total Words in Document",
    "Causal Terms (Total Occurrences)",
    "Noncausal Terms (Total Occurrences)",
    "Causal Terms (% of Total Words)",
    "Noncausal Terms (% of Total Words)",
    "Ratio (Causal:Noncausal)"
  ),
  Value = base::c(
    base::format(total_words, big.mark = ","),
    base::format(causal_terms_total_occurrences, big.mark = ","),
    base::format(noncausal_terms_total_occurrences, big.mark = ","),
    base::paste0(base::round(causal_percentage, 2), "%"),
    base::paste0(base::round(noncausal_percentage, 2), "%"),
    base::round(causal_percentage / noncausal_percentage, 2)
  )
)

# Create flextable
ft_summary <- flextable::flextable(summary_stats) |>
  flextable::set_header_labels(
    Metric = "Metric",
    Value = "Value"
  ) |>
  flextable::theme_booktabs() |>
  flextable::autofit() |>
  flextable::bold(j = 1) |>
  flextable::bg(i = ~ Metric %in% base::c("Causal Terms (% of Total Words)", 
                                           "Noncausal Terms (% of Total Words)"),
                bg = "#FFF9E6") |>
  flextable::align(j = 2, align = "right", part = "all")

ft_summary
```

# Detailed Term Counts

## Top Causal Terms

```{r}
#| label: top-causal-terms
#| tbl-cap: "Most Frequent Causal Terms"

# Filter to terms with at least 1 occurrence
causal_top <- causal_term_counts_df |>
  dplyr::filter(count > 0) |>
  dplyr::mutate(
    percentage = base::round((count / base::sum(count)) * 100, 1),
    cumulative_pct = base::round(base::cumsum(percentage), 1)
  )

# Create flextable for top 20 causal terms
ft_causal <- flextable::flextable(utils::head(causal_top, 20)) |>
  flextable::set_header_labels(
    term = "Causal Term",
    count = "Count",
    percentage = "% of Causal Terms",
    cumulative_pct = "Cumulative %"
  ) |>
  flextable::theme_booktabs() |>
  flextable::autofit() |>
  flextable::align(j = 2:4, align = "right", part = "all") |>
  flextable::bg(i = ~ count >= 10, bg = "#E8F4F8") |>
  flextable::bold(i = 1, part = "all")

ft_causal
```

## Top Noncausal Terms

```{r}
#| label: top-noncausal-terms
#| tbl-cap: "Most Frequent Noncausal Terms"

# Filter to terms with at least 1 occurrence
noncausal_top <- noncausal_term_counts_df |>
  dplyr::filter(count > 0) |>
  dplyr::mutate(
    percentage = base::round((count / base::sum(count)) * 100, 1),
    cumulative_pct = base::round(base::cumsum(percentage), 1)
  )

# Create flextable for top 20 noncausal terms
ft_noncausal <- flextable::flextable(utils::head(noncausal_top, 20)) |>
  flextable::set_header_labels(
    term = "Noncausal Term",
    count = "Count",
    percentage = "% of Noncausal Terms",
    cumulative_pct = "Cumulative %"
  ) |>
  flextable::theme_booktabs() |>
  flextable::autofit() |>
  flextable::align(j = 2:4, align = "right", part = "all") |>
  flextable::bg(i = ~ count >= 10, bg = "#FFF4E6") |>
  flextable::bold(i = 1, part = "all")

ft_noncausal
```

# Visualizations

## Comparison Bar Chart

```{r}
#| label: comparison-chart
#| fig-cap: "Causal vs. Noncausal Language Usage"
#| fig-width: 10
#| fig-height: 6

# Create comparison data
comparison_data <- base::data.frame(
  Category = base::c("Causal Terms", "Noncausal Terms"),
  Percentage = base::c(causal_percentage, noncausal_percentage),
  Count = base::c(causal_terms_total_occurrences, noncausal_terms_total_occurrences)
)

# Create bar chart
p_comparison <- ggplot2::ggplot(comparison_data, 
                                 ggplot2::aes(x = Category, y = Percentage, fill = Category)) +
  ggplot2::geom_col(width = 0.6, alpha = 0.8) +
  ggplot2::geom_text(ggplot2::aes(label = base::paste0(base::round(Percentage, 2), "%\n(n=", Count, ")")),
                     vjust = -0.5, size = 5, fontface = "bold") +
  ggplot2::scale_fill_manual(values = base::c("Causal Terms" = "#4A90E2", 
                                               "Noncausal Terms" = "#E89C4A")) +
  ggplot2::labs(
    title = "Percentage of Document Devoted to Causal vs. Noncausal Language",
    subtitle = base::paste("Total words in document:", base::format(total_words, big.mark = ",")),
    x = NULL,
    y = "Percentage of Total Words (%)"
  ) +
  ggplot2::theme_minimal(base_size = 14) +
  ggplot2::theme(
    legend.position = "none",
    plot.title = ggplot2::element_text(face = "bold", size = 16),
    plot.subtitle = ggplot2::element_text(size = 12),
    axis.text.x = ggplot2::element_text(face = "bold", size = 12)
  ) +
  ggplot2::ylim(0, base::max(comparison_data$Percentage) * 1.2)

p_comparison
```

## Top Terms Comparison

```{r}
#| label: top-terms-comparison
#| fig-cap: "Top 10 Most Frequent Terms by Category"
#| fig-width: 12
#| fig-height: 8

# Prepare top 10 from each category
top_causal <- causal_term_counts_df |>
  dplyr::filter(count > 0) |>
  utils::head(10) |>
  dplyr::mutate(category = "Causal")

top_noncausal <- noncausal_term_counts_df |>
  dplyr::filter(count > 0) |>
  utils::head(10) |>
  dplyr::mutate(category = "Noncausal")

# Combine
top_combined <- dplyr::bind_rows(top_causal, top_noncausal)

# Create faceted bar chart
p_top_terms <- ggplot2::ggplot(top_combined, 
                                ggplot2::aes(x = stats::reorder(term, count), y = count, fill = category)) +
  ggplot2::geom_col(alpha = 0.8) +
  ggplot2::geom_text(ggplot2::aes(label = count), hjust = -0.3, size = 3.5) +
  ggplot2::coord_flip() +
  ggplot2::facet_wrap(~ category, scales = "free_y") +
  ggplot2::scale_fill_manual(values = base::c("Causal" = "#4A90E2", 
                                               "Noncausal" = "#E89C4A")) +
  ggplot2::labs(
    title = "Top 10 Most Frequent Terms in Each Category",
    x = NULL,
    y = "Frequency Count"
  ) +
  ggplot2::theme_minimal(base_size = 12) +
  ggplot2::theme(
    legend.position = "none",
    plot.title = ggplot2::element_text(face = "bold", size = 14),
    strip.text = ggplot2::element_text(face = "bold", size = 12),
    panel.grid.major.y = ggplot2::element_blank()
  )

p_top_terms
```

# Complete Term Lists

## All Causal Terms (with occurrences)

```{r}
#| label: all-causal-terms
#| tbl-cap: "Complete List of Causal Terms Found in Document"

# Filter to only terms found in document
all_causal_found <- causal_term_counts_df |>
  dplyr::filter(count > 0)

ft_all_causal <- flextable::flextable(all_causal_found) |>
  flextable::set_header_labels(
    term = "Causal Term",
    count = "Occurrences"
  ) |>
  flextable::theme_booktabs() |>
  flextable::autofit() |>
  flextable::align(j = 2, align = "right", part = "all") |>
  flextable::bg(i = ~ count >= stats::median(count), bg = "#F0F8FF")

ft_all_causal
```

## All Noncausal Terms (with occurrences)

```{r}
#| label: all-noncausal-terms
#| tbl-cap: "Complete List of Noncausal Terms Found in Document"

# Filter to only terms found in document
all_noncausal_found <- noncausal_term_counts_df |>
  dplyr::filter(count > 0)

ft_all_noncausal <- flextable::flextable(all_noncausal_found) |>
  flextable::set_header_labels(
    term = "Noncausal Term",
    count = "Occurrences"
  ) |>
  flextable::theme_booktabs() |>
  flextable::autofit() |>
  flextable::align(j = 2, align = "right", part = "all") |>
  flextable::bg(i = ~ count >= stats::median(count), bg = "#FFFAF0")

ft_all_noncausal
```

# Interpretation

## Key Findings

Based on this analysis:

1. **Overall Language Balance**: The document uses **`r round(causal_percentage, 2)`%** causal language and **`r round(noncausal_percentage, 2)`%** noncausal language.

2. **Dominant Language Type**: `r if(causal_percentage > noncausal_percentage) "Causal language predominates" else "Noncausal language predominates"`, with a ratio of **`r round(causal_percentage / noncausal_percentage, 2)`:1**.

3. **Most Common Causal Term**: "**`r causal_term_counts_df$term[1]`**" appears **`r causal_term_counts_df$count[1]`** times.

4. **Most Common Noncausal Term**: "**`r noncausal_term_counts_df$term[1]`**" appears **`r noncausal_term_counts_df$count[1]`** times.

5. **Vocabulary Diversity**: 
   - **`r sum(causal_term_counts_df$count > 0)`** unique causal terms found
   - **`r sum(noncausal_term_counts_df$count > 0)`** unique noncausal terms found

## Implications

This analysis reveals the rhetorical stance of the document. A higher proportion of:

- **Causal language** suggests claims about cause-and-effect relationships, mechanisms, or explanatory frameworks
- **Noncausal language** suggests more descriptive, correlational, or associative claims

The balance between these language types can indicate whether the research emphasizes:
- Strong causal claims vs. more tentative associations
- Explanatory mechanisms vs. observed patterns
- Intervention effects vs. descriptive relationships

# Appendix

## Session Information

```{r}
#| label: session-info

utils::sessionInfo()
```

## File Paths Used

- **PDF Input**: `r pdf_input_path`
- **Causal Glossary**: `r causal_glossary_path`
- **Noncausal Glossary**: `r noncausal_glossary_path`
